---
title: "matching_external_cohort"
author: "Gabriel Innocenti"
date: "2026-02-27"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(vecmatch)
library(dplyr)

reference <- "simulated"
formula <- category ~ age * sex
```

## Define functions

```{r}
# Create database
generate_db <- function(metadata,
                 group_var = NULL, 
                 project_var = NULL, 
                 sex_var = NULL, 
                 age_interval = NULL) {
  
  # Remove NAs in sex and age
  metadata_vecmatch <- metadata %>%
    filter(!is.na(sex), !is.na(age))

  # Conditional filters
  if (!is.null(group_var)) {
    metadata_vecmatch <- metadata_vecmatch %>%
      filter(group_var == group_test)
  }

  if (!is.null(sex_var)) {
    metadata_vecmatch <- metadata_vecmatch %>%
      filter(sex_var == sex_var)
  }
  
  if (!is.null(project_var)) {
    metadata_vecmatch <- metadata_vecmatch %>%
      filter(project_var == project)
  }
  
  if (!is.null(age_interval) && length(age_interval) == 2) {
    metadata_vecmatch <- metadata_vecmatch %>%
      filter(age >= age_interval[1], age <= age_interval[2])
  }
  # Add the batch column (category)
  metadata_vecmatch["category"] <- "db"
  metadata_vecmatch <- metadata_vecmatch[c("sample_name","group_test", "sex", "age", "origin", "category")]
  return(metadata_vecmatch)
}


# Wrapper for running vecmatch using the metadata_db and giving the table with the expected distribution
run_vecmatch <- function(combined,
                         significance_raincloud = "t_test",
                         formula = NULL, # formula object
                         caliper = 1
) {
  formula = category ~ age * sex ### This should be possible to modify
  
  # 1. Calculate the Generalized propensity scores
  gps_matrix <- estimate_gps(formula,
                             data = combined,
                             method = "vglm",
                             reference = reference
  )
  
  # 2. Define the Common Support Region (CSR)
  csr_matrix <- csregion(gps_matrix)
  print(colnames(gps_matrix))
  str(gps_matrix)
  # 3. Matching on the generalized propensity scores
  matched_data <- match_gps(
    csmatrix = csr_matrix,
    reference = reference,
    caliper = caliper
  )
  return(list(matched_data, 
              csr_matrix, 
              gps_matrix))
}
```

## Load metadata slim and generate simulated data
```{r, echo=FALSE}
# Load metadata slim table
metadata <- read.table("metadata_slim.csv", sep=";", header=TRUE)
# Generate reference metadata table to be compared to a group defined by user
# In this case we want to select Controls from any project
# But we could select any disease present in the database, any specific project,
# specific age intervals and also specific project
metadata_db <- generate_db(metadata, 
                           group_var = "Controls") # Select based on disease group
summary(metadata_db)

```

## Simulate external data
Test first against simulated data on patients.
Here we define a population of 120 controls, age normally distributed with mean=60 and std=15, random sex sampling.

```{r}
set.seed(123)  # for reproducibility

# Parameters
n_samples <- 130           # total number of samples
age_mean <- 50             # average age
age_sd <- 15               # standard deviation for age

# Create simulated_data
simulated_data <- data.frame(
  sample_name = 1:n_samples,
  group_test = "Controls",
  sex = sample(c("F", "M"), n_samples, replace = TRUE),  # random F/M
  age = round(rnorm(n_samples, mean = age_mean, sd = age_sd)),  # normally distributed
  # age = round(runif(n_samples, min = 25, max = 90)), # randomly distributed in a given interval
  origin = "Austria",
  category = "simulated"
)

summary(simulated_data)
```
## Merge with the reference database (metadata_slim selected)
We merge the 2 groups and use vecmatch raincloud visualization to assess the sex-age distribution.
```{r}
# Concatenate the 2 tables
merged_data <- rbind(metadata_db, simulated_data)

# Use vecmatch visualization function to show the distribution of the 2 groups
raincloud(
  data = merged_data,
  y = age,
  group = category,
  facet=sex,
  significance = "t_test",
  sig_label_color = TRUE
)
```
Groups are heavily unmatched (sex,age). Standardized mean deviation is used as a metric to assess the difference between the 2 distribution and t-test for assessing the difference significance

## Use Vecmatch propensity score matching algoritm, using common support region calculation

```{r}
results <- run_vecmatch(combined =  merged_data)
matched_data <- results[[1]]
csr_matrix <- results[[2]]
gps_matrix <- results[[3]]

# Check distributions
raincloud(
  data = matched_data,
  y = age,
  group = category,
  facet=sex,
  significance = "t_test",
  sig_label_color = TRUE
)
```
Groups are now sex-age matched (non-significant difference).

## Remaining Issues

1. **External group cannot be treated as a fixed reference**

   In the current implementation, I could not identify a straightforward way to treat the external group (`simulated_data`) as a fixed reference group. The matching algorithm applies trimming to both groups: not only to the target samples (which may be out-of-distribution), but also to the reference group. This is problematic because the external samples are intended to preserve a distribution representative of the database and ideally should remain unaffected by trimming.

2. **Desirable two-step matching strategy**

   Ideally, I would implement a simplified two-step procedure. First, perform exact 1-to-1 matching within the database (i.e., for each external sample, identify an exact age–sex match in the internal database), thereby avoiding a full `vecmatch` run. Only if at least one exact match cannot be found should the algorithm fall back to generalized propensity score matching using `vecmatch`.
